---
title: "NYC Building Violation Analysis"
author: "Shreya Shetty (svs2148) & Shruti Shetty (ss7592)"
editor: visual
execute:
  echo: true
format:
  html:
    fig-width: 6
    fig-height: 4
    out-width: 60%
    embed-resources: true
---

## Loading Libraries

```{r}
# Libraries
library(dplyr)
library(lubridate)
library(janitor)
library(ggplot2)
library(ggalluvial)
library(tidyr)
library(scales)
library(data.table)
library(readr)
library(tidyverse)
library(forcats)
library(stringr)
library(redav)
library(GGally)
```

## Loading Housing Violations and 311 Complaints Datasets

```{r}
# ========== HOUSING VIOLATIONS ==========
housing_raw <- read.csv("/Users/shreyashetty/Documents/Fall 2025 Courses/EDAV/Final_Project/edav_project/datasets/Housing_Violations_2022_onwards.csv")

# ========== 311 HOUSING COMPLAINTS ==========
sr311_raw   <- read.csv("/Users/shreyashetty/Documents/Fall 2025 Courses/EDAV/Final_Project/edav_project/datasets/311_Housing_Complaints_2022_onwards.csv")

```

## Formatting dates and making sure loaded data has only Manhattan Borough entries

Here, we created a robust set of parse orders for mixed formats in both our datasets. We created cleaned date columns while keep original columns untouched. Also, we added new columns with month-year (Month name + Year) extracted which will be used for alluvial stacks later on

```{r}
# VALID DATE ORDERS

date_orders <- c("mdY HMS", "mdY HM", "mdY", "Ymd HMS", "Ymd")

housing <- housing_raw |>
  mutate(
    borough_std = toupper(Borough),
    # PARSE master versions — keep duplicates untouched
    inspectiondate_clean = parse_date_time(InspectionDate, orders = date_orders),
    approveddate_clean   = parse_date_time(ApprovedDate, orders = date_orders),
    certifieddate_clean  = parse_date_time(CertifiedDate, orders = date_orders),
    novissued_clean      = parse_date_time(NOVIssuedDate, orders = date_orders),
    statusdate_clean     = parse_date_time(CurrentStatusDate, orders = date_orders),
    originalcertifybydate_clean = parse_date_time(OriginalCertifyByDate, orders = date_orders),
    originalcorrectbydate_clean = parse_date_time(OriginalCorrectByDate, orders = date_orders),
    newcertifybydate_clean = parse_date_time(NewCertifyByDate, orders = date_orders),
    newcorrectbydate_cean = parse_date_time(NewCorrectByDate, orders = date_orders),
  ) |>
  filter(borough_std == "MANHATTAN")

sr311 <- sr311_raw |>
  mutate(
    borough_std = toupper(Borough),
    created_clean = parse_date_time(Created.Date, orders = date_orders),
    closed_clean  = parse_date_time(Closed.Date, orders = date_orders)
  ) |>
  filter(borough_std == "MANHATTAN")


# adding month-year (Month name + Year) columns used for alluvial stacks

housing <- housing |>
  mutate(
    inspection_month = ifelse(!is.na(inspectiondate_clean),
                              format(floor_date(inspectiondate_clean, "month"), "%B %Y"),
                              NA_character_),
    approved_month = ifelse(!is.na(approveddate_clean), 
                            format(floor_date(approveddate_clean,"month"), "%B %Y"),
                            NA_character_),
    nov_month = ifelse(!is.na(novissued_clean),
                       format(floor_date(novissued_clean, "month"), "%B %Y"),
                       NA_character_),
    status_month = ifelse(!is.na(statusdate_clean),
                          format(floor_date(statusdate_clean, "month"), "%B %Y"),
                          NA_character_)
  )

# Hosuing violations rows
nrow(housing)

sr311 <- sr311 |>
  mutate(
    created_month = ifelse(!is.na(created_clean),
                           format(floor_date(created_clean, "month"), "%B %Y"),
                           NA_character_),
    closed_month = ifelse(!is.na(closed_clean),
                           format(floor_date(closed_clean, "month"), "%B %Y"),
                           NA_character_)
)

# 311 requests rows updated
nrow(sr311)

```

The date columns have been parsed properly and as we can see the both raw data we downloaded, only had data from the Manhattan borough since before and after filtering the number of rows remain same for both datastes i.e. 589005 rows in Housing Violations dataset and 425248 rows in 311 requests dataset.

## Housing violation decription column data cleaning

```{r}
## Housing violation description column data cleaning

# First, let's inspect NOVDescription
head(housing$NOVDescription, 10)

# ========== REFINED CATEGORIZATION ==========
housing <- housing |>
  mutate(
    ViolationCategory = case_when(

      # 1 — Registration (Class I)
      Class == "I" |
        str_detect(NOVDescription, regex("registration|register", ignore_case = TRUE)) ~
        "REGISTRATION/ADMIN",

      # 2 — Smoke / CO
      str_detect(NOVDescription, regex("smoke detector|carbon monoxide|co detect", ignore_case = TRUE)) ~
        "SMOKE/CO DETECTOR",

      # 3 — Heat / Hot Water
      str_detect(NOVDescription, regex("heat|heating|hot water|radiator|boiler|steam|too cold", ignore_case = TRUE)) ~
        "HEAT/HOT WATER",

      # 4 — Plumbing
      str_detect(NOVDescription, regex("plumb|pipe|faucet|drain|sewer|bathroom|sink|toilet|water supply", ignore_case = TRUE)) ~
        "PLUMBING",

      # 5 — Water leak
      str_detect(NOVDescription, regex("leak|leaking|leakage|water drip", ignore_case = TRUE)) ~
        "WATER LEAK",

      # 6 — Paint / Plaster
      str_detect(NOVDescription, regex("paint|plaster|peel|wall|ceiling", ignore_case = TRUE)) ~
        "PAINT/PLASTER",

      # 7 — Door / Window / Lock
      str_detect(NOVDescription, regex("door|window|lock|self-closing|entrance", ignore_case = TRUE)) ~
        "DOOR/WINDOW/LOCK",

      # 8 — Pest / Sanitation
      str_detect(NOVDescription, regex("rodent|pest|roach|mice|rat|garbage|infest|sanitation", ignore_case = TRUE)) ~
        "PEST/SANITATION",

      # 9 — Floor / Ceiling
      str_detect(NOVDescription, regex("floor|tile|carpet|wood floor|ceramic", ignore_case = TRUE)) ~
        "FLOOR/CEILING",

      # 10 — Elevator
      str_detect(NOVDescription, regex("elevator|lift", ignore_case = TRUE)) ~
        "ELEVATOR",

      # 11 — Mold
      str_detect(NOVDescription, regex("mold|mildew|moisture", ignore_case = TRUE)) ~
        "MOLD",

      # 12 — Electrical
      str_detect(NOVDescription, regex("electric|wiring|outlet|light|circuit", ignore_case = TRUE))  ~ "ELECTRICAL",

      # 13 — Gas / Appliances
      str_detect(NOVDescription, regex("gas|appliance|stove", ignore_case = TRUE)) ~
        "GAS/APPLIANCES",

      # 14 — Ventilation
      str_detect(NOVDescription, regex("ventilat|airflow|exhaust fan", ignore_case = TRUE)) ~
        "VENTILATION",

      # 15— Fire Safety
      str_detect(NOVDescription, regex("fire|sprinkler|fire escape|extinguish", ignore_case = TRUE)) ~
        "FIRE SAFETY",

      # 16— Intercom / Bell
      str_detect(NOVDescription, regex("bell|buzzer|intercom", ignore_case = TRUE)) ~
        "BUILDING SYSTEMS",

      # 17 — Building Management
      str_detect(NOVDescription, regex("janitor|superintendent|building service|super", ignore_case = TRUE)) ~
        "BUILDING MANAGEMENT",

      # Default
      TRUE ~ "OTHER"
    )
  )

# Check distribution of newly created categories
housing |>
  count(ViolationCategory, sort = TRUE) |>
  print()
```

```{r}
# ========== 311: CATEGORY MAPPING (parallel to housing) ==========

# Compare with 311 Complaint Types
sr311 |>
  count(Complaint.Type, sort = TRUE) |>
  print()

# Mapped categories
sr311 <- sr311 |>
  mutate(
    Category = case_when(
      # Heat / Hot Water
      str_detect(`Complaint.Type`, regex("HEAT|HOT WATER", ignore_case = TRUE)) ~ "HEAT/HOT WATER",
      
      # Plumbing
      str_detect(`Complaint.Type`, regex("PLUMBING", ignore_case = TRUE)) ~ "PLUMBING",
      
      # Water Leak
      str_detect(`Complaint.Type`, regex("WATER LEAK", ignore_case = TRUE)) ~ "WATER LEAK",
      
      # Elevator
      str_detect(`Complaint.Type`, regex("ELEVATOR", ignore_case = TRUE)) ~ "ELEVATOR",
      
      # Mold
      str_detect(`Complaint.Type`, regex("MOLD", ignore_case = TRUE)) ~ "MOLD",
      
      # PAINT / PLASTER — newly added
      str_detect(`Complaint.Type`, regex("PAINT|PLASTER", ignore_case = TRUE)) ~ "PAINT/PLASTER",
      
      # Pest / sanitation based on Complaint.Type AND Descriptor
      (
        str_detect(`Complaint.Type`, regex("MAINTENANCE|FACILITY", ignore_case = TRUE)) &
        str_detect(Descriptor, regex("rodent|rodents|mice|rats|mouse|rat|insect|pest", ignore_case = TRUE))
      ) ~ "PEST/SANITATION",
      
      # Fallback
      TRUE ~ "OTHER"
    )
  )

# New mapped categories
sr311 |>
  count(Category, sort = TRUE) |>
  print()

```

Here, manual categorization logic we applied successfully maps unstructured text descriptions (NOVDescriptions) in Housing Violations dataset into unified high-level categories which enables a direct one-to-one comparison between both. We also normalized the categories in 311 Complaints dataset since there were multiple for the same category, ege. there were 2 types "HEAT/HOT WATER" and "Heat/Hot Water" which were merged into 1.

## Save cleaned CSV copies for reproducibility

```{r}
write.csv(housing, "datasets/housing_manhattan_3years_clean.csv", row.names = FALSE)
write.csv(sr311,   "datasets/sr311_manhattan_3years_clean.csv", row.names = FALSE)
```

## CATEGORY COMPARISON: Housing vs 311

```{r}
#| fig-height: 7
comparison_categories <- c(
  "HEAT/HOT WATER", "PLUMBING", "PAINT/PLASTER",
  "WATER LEAK", "ELEVATOR", "MOLD", "PEST/SANITATION"
)

# Violations count
viol_comparison <- housing |>
  filter(ViolationCategory %in% comparison_categories) |>
    group_by(ViolationCategory) |>
      summarise(Count = n(), .groups = "drop") |>
        mutate(
          Category = ViolationCategory,
          Source = "Housing Violations"
        ) |>
          select(Category, Count, Source)

# 311 count
complaints_comparison <- sr311 |>
  filter(Category %in% comparison_categories) |>
    group_by(Category) |>
      summarise(Count = n(), .groups = "drop") |>
        mutate(Source = "311 Complaints")

# Combine both
combined <- bind_rows(viol_comparison, complaints_comparison)

# Plot
ggplot(combined, aes(x = reorder(Category, -Count), y = Count, fill = Source)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = scales::comma(Count)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c(
    "Housing Violations" = "purple",
    "311 Complaints"     = "pink"
  )) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Violations vs 311 Complaints by Category",
    subtitle = "Manhattan (Last 3 yrs) Comparable Categories Only",
    x = "Category",
    y = "Count",
    fill = "Source"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 17),
    plot.subtitle = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

```

There is a substantial disparity in volume for the "HEAT/HOT WATER" category in violations vs complaints, where the number of 311 Service Requests is significantly higher than the count of official Housing violations issued for the same. Hence, we can conclude that while tenant complaints regarding heating are very very frequent, quite a smaller proportion result in confirmed enforcement actions by management. In other categories, the difference is relatively small. In case of Paint/Plaster and Pests, there is much higher number of building violation entries which shows that probably regular building inspections is conducted for those and hence less direct 311 complaints. Out of all top 7 identified categories overlapping in both the datasets, Heat/Hot Water is the highest and Mold is the lowest complaint type.

## 3-year per-quarter faceted plots

```{r}
#| fig-height: 7
# ----- HOUSING: QUARTERLY CATEGORY TRENDS (3-LINE-QTRS) -----

housing_quarter_lines <- housing |>
  filter(!is.na(inspectiondate_clean)) |>
    mutate(
      year = year(inspectiondate_clean),
      quarter = paste0("Q", quarter(inspectiondate_clean))
    ) |>
      count(year, quarter, ViolationCategory) |>
        mutate(
          quarter = factor(quarter, levels = c("Q1","Q2","Q3","Q4"))
        )

focus_cats <- c(
  "HEAT/HOT WATER",
  "PLUMBING",
  "PAINT/PLASTER",
  "WATER LEAK",
  "ELEVATOR",
  "MOLD",
  "PEST/SANITATION"
)

housing_quarter_lines_focus <- housing_quarter_lines |>
  dplyr::filter(ViolationCategory %in% focus_cats)

ggplot(housing_quarter_lines_focus,
       aes(x = quarter, y = n, color = factor(year), group = year)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("2023" = "#E63946",
                                "2024" = "#457B9D",
                                "2025" = "#2A9D8F"),
                     name = "Year") +
  facet_wrap(~ ViolationCategory,
             scales = "free_y",
             ncol = 2) +    
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 17),
    axis.text.x = element_text(angle = 0),
    legend.position = "top"
  ) +
  labs(
    title = "Quarterly Housing Violations by Category",
    subtitle = "Manhattan (2022–25) • Selected Categories",
    x = "Quarter",
    y = "Number of Violations"
  )


# ----- 311: QUARTERLY CATEGORY TRENDS (3-LINE-QTRS) -----

sr311_quarter_lines <- sr311 |>
  filter(!is.na(created_clean)) |>
  mutate(
    year    = year(created_clean),
    quarter = paste0("Q", quarter(created_clean))
  ) |>
  count(year, quarter, Category) |>
  mutate(
    quarter = factor(quarter, levels = c("Q1","Q2","Q3","Q4"))
  )

sr311_quarter_lines_focus <- sr311_quarter_lines |>
  filter(Category %in% focus_cats)

ggplot(sr311_quarter_lines_focus,
       aes(x = quarter, y = n, color = factor(year), group = year)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("2023" = "#E63946",
                                "2024" = "#457B9D",
                                "2025" = "#2A9D8F"),
                     name = "Year") +
  facet_wrap(~ Category, scales = "free_y", ncol = 2) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title  = element_text(face = "bold", size = 17),
    axis.text.x = element_text(angle = 0),
    legend.position = "top"
  ) +
  labs(
    title    = "Quarterly 311 Complaints by Category",
    subtitle = "Manhattan (2022-25) • Selected Categories",
    x = "Quarter",
    y = "Number of Complaints"
  )

```

Across quarters, HEAT/HOT WATER 311 complaints and housing violations both show a clear seasonal pattern. They spike in Q1, drop sharply in Q2 and Q3 for all 3 years, and rise again Q4 matching winter heating needs. For PLUMBING, PAINT/PLASTER, PEST/SANITATION, WATER LEAK, ELEVATOR, and MOLD, housing violations counts change over the quarters (most show a local peak around Q3), but patterns are less sharp and not perfectly identical so they don't show a single clear seasonal shape like HEAT/HOT WATER does. Pest and sanitation 311 complaints peak in Q3 in all 3 years, which suggests these problems are most common in late summer, which is also consistent with warmer weather since heat and humidity make it easier for rodents, roaches, and insects to spread, and garbage issues can get worse when it’s so hot. But it's not as strong or as perfectly consistent across all years. For the other categories for 311 complaints, the quarter‑to‑quarter changes within each panel are relatively small and do not follow a clear, repeated seasonal shape across years.

## Housing Violation Monthly Transition Alluvial for 2024 (for a small sample size)

```{r}
# ---- Housing Violation 2024 Monthly Transition Alluvial ----

set.seed(200)
month_levels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

viol_flow_2024 <- housing |>
  select(ViolationID, Class,
         inspectiondate_clean, novissued_clean, statusdate_clean) |>
  mutate(
    insp_m = floor_date(inspectiondate_clean, "month"),
    nov_m  = floor_date(novissued_clean,       "month"),
    stat_m = floor_date(statusdate_clean,      "month")
  ) |>
  filter(!is.na(insp_m),
         year(insp_m) == 2024,
         (!is.na(nov_m) | !is.na(stat_m))) |>
  slice_sample(n = 200) |>
  mutate(
    insp_m_lab = factor(format(insp_m, "%b"),  levels = month_levels),
    nov_m_lab  = factor(format(nov_m,  "%b"),  levels = month_levels),
    stat_m_lab = factor(format(stat_m, "%b"),  levels = month_levels)
  ) |>
  group_by(insp_m_lab, nov_m_lab, stat_m_lab, Class) |>
  summarise(n = n(), .groups = "drop")


# plot
ggplot(viol_flow_2024,
       aes(axis1 = insp_m_lab, axis2 = nov_m_lab, axis3 = stat_m_lab, y = n)) +
  geom_alluvium(aes(fill = Class),
                width = 0.3, alpha = 0.9) +   # softer, smoother ribbons
  geom_stratum(aes(fill = Class),
               width = 0.3, alpha = 0.2) +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum)),
            size = 3, color = "black") +
  scale_fill_brewer(type=qual, palette = "Set2") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "2024 Housing Violations Flow: Inspection → NOV → Status (Sample)",
    x = "Stage (Month in 2024)",
    y = "Number of violations"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 11),
    legend.title = element_text(face = "bold")
  )

```

We can see how 2024 housing violations (we have taken a small sample slice of 200 here to visualize better) move from inspection month to NOV (notivr of violation) month to when current status takes effect, with ribbon colors showing violation Class (A least serious, C most serious, I immediately hazardous). Mostly, flows stay within nearby months across 3 stages showing many violations progress through the process within the same part of the year. Thicker orange and blue ribbons show that Class B and C violations make up most of the pipeline, while Class A is thinner and Class I appears as a smaller set of immediately hazardous cases. There are also some blank bands in NOV step which occur when inspection and status dates are recorded but NOVIssuedDate or Class code is missing, so those months or classes appear as NA.

## Time Series

### Housing Daily

```{r}
housing |>
  filter(!is.na(inspectiondate_clean)) |>
  count(inspectiondate_clean) |>
  ggplot(aes(x = inspectiondate_clean, y = n)) +
  geom_line(color = "purple") +
  labs(title = "Housing Violations Timeline (daily, Manhattan, 2022-2025)",
  x = "Inspection date", y = "Count") +
  theme_minimal()
```
The spikes here kind of look like a pattern, but the size and shape suggest they are just random glitches and not exactly a real trend. On normal days between 2022-2025, it is mostly noisy and a flat band around a few hundred inspections but on very few days it jumps over 5,000. This is extremely high & not very realistic just 1 day and does not spread over nearby days so it's not a trend but these spikes are outliers.

### 311 Daily

```{r}
sr311 |>
  filter(!is.na(created_clean)) |>
  count(created_clean) |>
  ggplot(aes(x = created_clean, y = n)) +
  geom_line(color = "pink") +
  labs(title = "311 Requests Timeline (daily, Manhattan, 2022-2025)",
  x = "Created date", y = "Count") +
  theme_minimal()
```

It is almost a solid pink block with very tiny variation in consecutive days and that's why it's very hard to see any clear pattern. Hence, this plot does no give us much information. TO further deep dive into finding a trend in the time series, we plotted combined monthly series for both the dataset.
 
### Monthly combined of housing and 311

```{r}
housing_month <- housing |>
  filter(!is.na(inspectiondate_clean)) |>
  mutate(month = floor_date(inspectiondate_clean, "month")) |>
  count(month, name = "HousingViolationCount")

sr311_month <- sr311 |>
  filter(!is.na(created_clean)) |>
  mutate(month = floor_date(created_clean, "month")) |>
  count(month, name = "311ComplaintCount")

monthly_combined <- full_join(housing_month, sr311_month, by = "month") |>
  replace_na(list(HousingCount = 0, ComplaintCount = 0))

monthly_combined |>
  pivot_longer(cols = c("HousingViolationCount", "311ComplaintCount"), names_to = "Source", values_to = "Count") |>
  ggplot(aes(x = month, y = Count, color = Source)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("311ComplaintCount" = "pink", "HousingViolationCount" = "purple"))+
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = , size = 5)) +
  labs(title = "Monthly: Housing Violations vs 311 Complaints (Manhattan)",
  x = "Month", y = "Count")
```





Here, both 311 complaints and housing violations in Manhattan during 2022-2025 have a fairly strong seasonal pattern, but there are also some exceptions. In most years the red complaint line is highest in late fall & winter, & lowest in late spring & summer which basically means people call 311 line more about housing problems during colder months when heat and building conditions matter more. Now  blue violations line also have major winter peaks overall but it includes few extra spikes in other months (like June & August 2025), so its pattern is winter‑dominated but not specific to winter‑only. Also, some of those spikes might be anomalies and outlier inspections instead of a reasonable trend.

### Time series Variation by Month and Year

```{r}


set.seed(2025)

# Load data
housing_2 <- as.data.table(housing)

# Convert date and extract time components
housing_2[, InspectionDate := as.Date(InspectionDate)]
housing_2[, YearMonth := format(InspectionDate, "%Y-%m")]
housing_2[, Month := month(InspectionDate)]
housing_2[, Year := year(InspectionDate)]

# Count violations by month and year
viol_by_month <- housing_2[, .N, by = .(Year, Month)][order(Year, Month)]

# Create the plot
ggplot(viol_by_month, aes(x = Month, y = N, color = factor(Year), group = Year)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = 1:12, 
                     labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Housing Violations by Month and Year",
    subtitle = "Manhattan, 2022-2025",
    x = "Month",
    y = "Number of Violations",
    color = "Year"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )




```

It shows that there is no stable seasonal pattern across years. Some months are high in one year and low in another like February is very unsually high only in 2022, while peaks in other years appear in different months like April 2024 or June 2025. Hence, we can conclude that monthly counts fluctuate randomly rather than following a consistent winter–summer or any other repeating trend.

### Investingating February 2022 peak

```{r}

# Investigate February 2022
feb_2022 <- housing_2[Year == 2022 & Month == 2]
cat("February 2022 violations:", nrow(feb_2022), "\n")

# Check if they're all inspected in Feb or entered in Feb
summary(feb_2022$InspectionDate)
summary(feb_2022$ApprovedDate)


```
Upon further review we can see that Median/Mean inspection date is February 15, 2022 & most inspections are clustered around Feb 15-16 only. This does not seem like a data error but is simply  appearing to be a mass inspection event in mid-February 2022. It is not a trend.


### Smoothed Temporal View

We tried to find any overall trends when we smooth out the monthly fluctuations

```{r}

# Aggregate by quarter for smoother trend
housing_2[, Quarter := quarter(InspectionDate)]
housing_2[, YearQuarter := paste0(Year, "-Q", Quarter)]

viol_by_quarter <- housing_2[, .N, by = .(Year, Quarter, YearQuarter)][order(Year, Quarter)]

# Create quarter labels
viol_by_quarter[, QuarterLabel := paste0(Year, "\nQ", Quarter)]

# Plot
ggplot(viol_by_quarter, aes(x = factor(YearQuarter, levels = YearQuarter), 
                              y = N, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Housing Violations by Quarter",
    subtitle = "Manhattan, 2022-2025 (smoothed trend)",
    x = "Quarter",
    y = "Number of Violations"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )



```

It shows that housing violations in Manhattan from 2022-2025 move up and down between quarters but stay in a fairly similar range (30,000–45,000), so there is no strong upward or downward trend. Q4 20205 seems like an anomaly but it is not because it does not include data from mid November - December 2025 data are not fully included, hence it is not any abnormal evidence of a real drop.




## Spatial sample plots

### Housing Violations

```{r}
housing |>
  filter(!is.na(Longitude) & !is.na(Latitude)) |>
  slice_sample(n = 5000) |>
  ggplot(aes(x = Longitude, y = Latitude)) +
  geom_point(alpha = 0.4, size = 0.6) +
  theme_minimal() +
  labs(title = "Housing Violations Locations (sample 5k)", x = "Longitude", y = "Latitude")
```

### 311 Complaints

```{r}
sr311 |>
  filter(!is.na(Longitude) & !is.na(Latitude)) |>
  slice_sample(n = 5000) |>
  ggplot(aes(x = Longitude, y = Latitude)) +
  geom_point(alpha = 0.4, size = 0.6, color = "orange") +
  theme_minimal() +
  labs(title = "311 Complaints Locations (sample 5k)", x = "Longitude", y = "Latitude")
```



## Parallel Coordinated plot

```{r}

#| fig-height: 8

housing_dt <- housing_2
sr311_dt   <- as.data.table(sr311)

housing_dt[, BBL := as.character(BBL)]
sr311_dt[,  BBL := as.character(BBL)]

viol_severity <- housing_dt[
  !is.na(BBL) & BBL != "",
  .(
    ViolationCount = .N,
    ClassC         = sum(Class == "C"),
    RentImp        = sum(RentImpairing == "Y")
  ),
  by = BBL
]

viol_severity[, Pct_ClassC  := ClassC  / ViolationCount]
viol_severity[, Pct_RentImp := RentImp / ViolationCount]

complaints_by_building <- sr311_dt[
  !is.na(BBL) & BBL != "",
  .(ComplaintCount = .N),
  by = BBL
]

bldg_pcp <- merge(
  viol_severity,
  complaints_by_building,
  by = "BBL",
  all.x = TRUE
)
bldg_pcp[is.na(ComplaintCount), ComplaintCount := 0]

bldg_pcp_filt <- bldg_pcp[
  ViolationCount >= 20 | ComplaintCount >= 50
]

pcp_df <- as.data.frame(bldg_pcp_filt[, .(
  ViolationCount = as.numeric(ViolationCount),
  ComplaintCount = as.numeric(ComplaintCount),
  Pct_ClassC     = as.numeric(Pct_ClassC),
  Pct_RentImp    = as.numeric(Pct_RentImp)
)])

set.seed(2025)

# Sample 300 buildings
pcp_sample <- pcp_df[sample(nrow(pcp_df), 300), ]

# Create severity groups based on Class C share
pcp_sample$SeverityGroup <- cut(
  pcp_sample$Pct_ClassC,
  breaks = c(-Inf, 0.2, 0.5, Inf),
  labels = c("Low C share", "Medium C share", "High C share")
)


ggparcoord(
  data        = pcp_sample,
  columns     = 1:4,
  groupColumn = "SeverityGroup",
  scale       = "uniminmax",
  alphaLines  = 0.25,          # lighter, smoother
  showPoints  = FALSE
) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 11) +   # smaller base font
  theme(
    plot.title   = element_text(face = "bold", size = 16),
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11),
    axis.text.x  = element_text(size = 9),   # reduce x‑axis text size
    axis.text.y  = element_text(size = 9),
    legend.position = "bottom",              # legend at bottom
    legend.title    = element_text(size = 10),
    legend.text     = element_text(size = 9)
  ) +
  labs(
    title = "Parallel Coordinates: Sample of Buildings",
    x = "Variables",
    y = "Scaled value",
    color = "Class C share"
  )
```

## Facet Comparative Panel (Housing vs 311)

```{r}
#| fig-height: 8
# ----------------------------
# 1) STANDARDIZE ZIPCODE NAME
# ----------------------------

housing2 <- housing |>
  rename(ZipCode = Postcode)

sr311_2 <- sr311 |>
  rename(ZipCode = Incident.Zip)

# ----------------------------
# 2) CREATE TOP SETS PER VARIABLE (20 each)
# ----------------------------

# Top 10 ZIPs
housing_zip <- housing2 |> filter(!is.na(ZipCode)) |> count(ZipCode) |> slice_max(n, n = 10)
sr311_zip   <- sr311_2 |> filter(!is.na(ZipCode)) |> count(ZipCode) |> slice_max(n, n = 10)

# Top 10 Streets
housing_st <- housing2 |> filter(!is.na(StreetName)) |> count(StreetName) |> slice_max(n, n = 10)
sr311_st   <- sr311_2 |> filter(!is.na(Street.Name)) |> count(Street.Name) |> slice_max(n, n = 10)

# Top Categories (Housing only but we compare counts against 311 Categories
housing_cat <- housing2 |> filter(!is.na(ViolationCategory)) |> count(ViolationCategory) |> slice_max(n, n = 10) |> rename(Category = ViolationCategory)
sr311_cat   <- sr311_2 |> filter(!is.na(Category)) |> count(Category) |> slice_max(n, n = 10) 

# Top 10 Blocks
housing_block<- housing2 |> filter(!is.na(Block)) |> count(Block) |> slice_max(n, n = 10)
sr311_block  <- sr311_2 |> filter(!is.na(BBL)) |> count(BBL) |> slice_max(n, n = 10)


# ----------------------------
# 3) COMBINE PANELS  (FIXED WITH as.character)
# ----------------------------

zip_panel <- bind_rows(
  housing_zip |> mutate(Value = as.character(ZipCode), Source = "Housing", Facet = "ZipCode"),
  sr311_zip   |> mutate(Value = as.character(ZipCode), Source = "311",     Facet = "ZipCode")
)


category_panel <- bind_rows(
  housing_cat |> mutate(Value = as.character(Category), Source = "Housing", Facet = "Category"),
  sr311_cat   |> mutate(Value = as.character(Category), Source = "311",     Facet = "Category")
)

street_panel <- bind_rows(
  housing_st |> mutate(Value = as.character(StreetName), Source = "Housing", Facet = "StreetName"),
  sr311_st   |> mutate(Value = as.character(Street.Name), Source = "311",     Facet = "StreetName")
)


block_panel <- bind_rows(
  housing_block |> mutate(Value = as.character(Block), Source = "Housing", Facet = "Block"),
  sr311_block  |> mutate(Value = as.character(BBL), Source = "311",     Facet = "Block")
)

# ----------------------------
# 4) FINAL COMBINED PANEL
# ----------------------------

panel_all <- bind_rows(zip_panel, street_panel, category_panel, block_panel) |>
  group_by(Facet) |>
  mutate(Value = fct_reorder(Value, n)) |>
  ungroup()# ----------------------------
# 4) FINAL FACETED PLOT
# ----------------------------

ggplot(panel_all, aes(x = Value, y = n, fill = Source)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~ Facet, scales = "free", ncol = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Housing" = "purple", "311" = "pink")) +
  labs(
    title = "Top Manhattan Characteristics Across Housing Violations and 311 Complaints",
    subtitle = "ZIP Codes, Street Names, Violation Categories, and Blocks",
    x = "",
    y = "Count",
    fill = "Source"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 18),
    legend.position = "top"
  )

```



## Top Violation Classes

What Types of Violations Are Most Common? Research Question: What are the most common housing violations that tenants and inspectors encounter?


```{r}
# ===  Violation Class Distribution ===
# Class definitions: A=Non-Hazardous, B=Hazardous, C=Immediately Hazardous, I=Failure to Register

class_data <- housing |>
  filter(!is.na(Class)) |>
  count(Class, name = "N") |>
  mutate(
    ClassLabel = case_when(
      Class == "A" ~ "Class A: Non-Hazardous",
      Class == "B" ~ "Class B: Hazardous",
      Class == "C" ~ "Class C: Immediately Hazardous",
      Class == "I" ~ "Class I: Failure to Register",
      TRUE ~ Class
    )
  ) |>
  arrange(desc(N))

ggplot(class_data, aes(x = reorder(ClassLabel, N), y = N, fill = Class)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::comma(N)), hjust = -0.1, size = 5) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Housing Violations by Severity Class",
    subtitle = "Manhattan, 2022-2025",
    x = "Violation Class",
    y = "Number of Violations"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

```

## Rent-Impairing Violation
```{r}
# === Rent-Impairing Violations ===

rent_data <- housing |>
  filter(!is.na(RentImpairing)) |>
  count(RentImpairing, name = "N") |>
  mutate(
    Label = ifelse(
      RentImpairing == "Y",
      "Rent-Impairing\n(affects habitability)",
      "Non-Rent-Impairing"
    )
  )

ggplot(rent_data, aes(x = Label, y = N, fill = RentImpairing)) +
  geom_col() +
  geom_text(
    aes(label = paste0(scales::comma(N), "\n(", round(N / sum(N) * 100, 1), "%)")),
    vjust = -0.5,
    size = 5
  ) +
  scale_fill_manual(values = c("N" = "steelblue", "Y" = "darkred")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Rent-Impairing vs Non-Rent-Impairing Violations",
    subtitle = "Manhattan, 2022-2025",
    x = "",
    y = "Number of Violations"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

```



## Save cleaned RDS as backups

```{r}
saveRDS(housing, file = "housing_manhattan_3years_clean.rds")
saveRDS(sr311,   file = "sr311_manhattan_3years_clean.rds")
```
